A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-13 Simulation/Writeup/NoteConvergence.Rnw
prelim conclusion

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.

@@

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
\textbf{say something about Rh below 1.}
%%%%

@@

\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}

@@

\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.

>>>>

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition. \textbf{add something about Rh below 1.}

@@

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
%%%%

@@

\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally ok. The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more. This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE?? COMPUTE??]. harder problems require more iterations.} %give some implications for these measures. connect convergence and simulation diagnostics.}

There is a discrepancy between what the convergence diagnostics and the performance measures indicate. While $T=4$ is ok with repsect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.

@@

\textbf{Answer RQ.} How to diagnose non-convergence? It's complicated.
%We will evaluate how convergence of the imputation algorithm can be diagnosed.

### 2019-12-13 Simulation/Writeup/NoteConvergence.tex
prelim conclusion

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition.

@@

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
\textbf{say something about Rh below 1.}
%%%%

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. At $T=5$, we reach the same $AC$-value as observed at $T=2$. $AC$-values plateau around $T=70$. $AC$ is then indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

@@

%% --------------------------------
\subsection{Simulation diagnostics}
\textbf{Bias.} Figure \ref{fig:sim} shows the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.

\textbf{Confidence interval width.} CIW is only clearly divergent for the simulation condition with one iteration. Conditions with two or more iterations have similar CIWs. The magnitude of differences from the loess line is somewhat larger up-to the condition with four iterations. From four iterations on, CIW seems to be stable across iterations. \textbf{Loess line is never flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Coverage rate.} \textbf{The coverage rate is more or less stable from two iterations upwards. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%. Loess line is never flat.}

\textbf{give some implications for these measures. connect convergence and simulation diagnostics.}

>>>>

Figure \ref{fig:conv}A shows that $\widehat{R}$-values generally decrease with increasing imputation chain lengths. The decline stabilizes somewhere between the simulation conditions $T=30$ and $T=50$. The downward trend is most pronounced for $T=3$, and between $T = 5$ and $T = 10$. In the intervening conditions ($3 \leq T \leq 5$), however, we observe a steep increase in $\widehat{R}$-values. This increase implies that convergence cannot be diagnosed at $T=3$, as would be the case based on the conventional threshold $\widehat{R} < 1.2$. \textbf{This may be spurious.} According to the widely used threshold $\widehat{R} < 1.1$, convergence can be diagnosed for conditions where $T>9$. If we use the recently recommended threshold $\widehat{R} < 1.01$, we would conclude that we cannot diagnose convergence in any simulation condition. \textbf{add something about Rh below 1.}

@@

%%%%
%Figure \ref{fig:conv} shows an initial dip, a small bump upwards, and then a steep decline. This decline continues up-to iteration twenty, after which a more gradual decrease can be observed up-to iteration forty. The $\widehat{R}$-values are more or less stable after that. 
%\textbf{[Remove stuff here]} %The conventionally acceptable threshold for convergence, maximum 
%$\widehat{R} < 1.2$ is reached at $T =3$, and conditions with six iterations or more. The widely used threshold $\widehat{R} < 1.1$ is met in conditions with ten or more iterations. The most recent recommended threshold is not reached within the 100 iterations considered in this simulation study. $\widehat{R} > 1.01$ for all simulation conditions (see online appendix for the full table of results).
%%%%

The $AC$-values displayed in Figure \ref{fig:conv}B are almost uniformly increasing as a function of $T$. The only $AC$-value that deviates from this observation is for $T=2$. The lowest $AC$-value is obtained for $T=3$. At $T=5$, the $AC$-value reaches the level observed at $T=2$. Between $T=10$ and $T=30$ the increase plateaus off, and $AC$-values are stable around $T=70$. $AC$ is then indifferentiable from zero, indicating stationarity. %Simulation conditions where $T>5$ have $AC$-values greater than $T=2$. %We see an initial decrease between $T=2$ and $T=3$.
According to this diagnostic, none of the simulation conditions show signs of non-convergence, since we only observe negative or zero $AC$-values.

@@

Taken together, we see that convergence may be diagnosed in simulation conditions where $T>10$.

%% --------------------------------
\subsection{Simulation diagnostics}
We use average bias, average condidence interval width, and coverage rate as performance measures to evaluate $\widehat{R}$ and $AC$. We make the general observation that the simulation diagnostics behave as theorized for most simulation conditions (bias around zero, coverage rate around 95\%).

Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations. Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.  
%the average bias between the regression coefficient of $X$ on $Y$ in the complete dataset, and $\bar{Q}$, \textbf{the pooled estimated regression coefficient and the ... Average bias is close to zero for simulation conditions with two iterations or more. The simulation condition with one iteration has a negative bias. From two iterations upwards, the average bias across repetitions is stable. Across iterations two to one hundred, bias fluctuates within a narrow range around zero. Loess line is flat around it = 20.} 
% Average bias is not zero because there is only one incomplete dataset to be imputed. Bias is a sample effect.
%Figure \ref{fig:sim}A shows that bias is fairly stable across simulation conditions. The condition $T=1$ clearly deviates from this trend with a negative bias. The bias for $T=2$ is below average, but within the range of fluctuations.  Conditions where $T>3$ \textbf{are ok, Loess line is flat for $T>20$}.

CIW is only clearly divergent for the simulation condition $T=1$. However, we want to be on the safe side, and not under-estimate the variance. Otherwise, we might end up with spurious inferences. Conditions where $T>3$ have similar CIWs, which seems to be stable across iterations. \textbf{Loess line is never completely flat.}%One might argue that there is some downward trending up-to thirthy to forty iterations. %After about twenty-five iterations the average CIW across all simulation conditions and repetitions is reached 0.928.

\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$. We see some over-coverage at $T=2$. We might over-estimate the variance, but better than under-estimating it. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely 95.3\%. Loess line is never flat.}

\textbf{From the simulation diagnostics, we see that simulation conditions where $T>3$ are generally ok. The performance measures indicate that for the current data, MI procedure, and quantity of interest, the algorithm should have maxit argument of four or more. This is, however, a simple simulation problem, with MCAR missingness mechanism, univariate missingness, and moderate fmi's [REMOVE?? COMPUTE??]. harder problems require more iterations.} %give some implications for these measures. connect convergence and simulation diagnostics.}

There is a discrepancy between what the convergence diagnostics and the performance measures indicate. While $T=4$ is ok with repsect to simulation quantities, $T=3$ is the worst possible value according to the convergence diagnostics, and $T=4$ provides marginally better convergence diagnostics.
